---
title: "Fisher's Exact Test"
author: "Tripp Bishop"
format: html
---

```{r setup}
#| echo: false
#| message: false

library(tidyverse)
library(effectsize)
library(pwr)
theme_set(theme_minimal())

gamedata <- read_csv("data/cookie_cats_subset.csv")
subgame <- gamedata |> 
  filter(sum_gamerounds >= 545)
```


Fisher's Exact test is another non-parametric test that can be used to test for relationships between groups in data. It is a good alternative to the Chi-square test when

* The sample size is < 1000 observations
* The frequency is < 5 for 20% or more of the group comparisons

## Effect size
The effect size for Fisher's Exact test is the odds ratio of the groups. 

* Small: 1.5
* Medium: 2.5
* Large: 4

The odds ratio is reported by `fisher.test` in R.

## Considerations
Fisher's Exact test is computationally expensive so if there are many groups in the data, it may be prohibitive to run. The advantage of Fisher's over Chi-square is that it provides an exact *p*-value rather than an approximations like Chi-square does.

## Examples

Fisher's Exact test is easy in R.
```{r}
freqtbl <- table(subgame$version, subgame$retention_7)
fisher.test(freqtbl)
```

The odds ratio in the output is used as the effect size. To compute the power of the test, the `pwr.2p.test` function can be used.

```{r}
pwr.2p.test(h=1.2535,
		   n=331,
		   sig.level=0.05)
```


