---
title: "Mann-Whitney U Test"
author: "Tripp Bishop"
format: html
---

```{r setup}
#| echo: false
#| message: false

library(tidyverse)
library(effectsize)
library(pwr)
theme_set(theme_minimal())

gamedata <- read_csv("data/cookie_cats_subset.csv")
```

The Mann-Whitney U test is a non-parametric test. It does not assume that the data have a specific distribution. This test is the non-parametric version of the independent t-test.

## Overview of the test
The Mann-Whitney U Test makes the following makes the assumption that the two samples have the same distribution shape (ex. both bell-shaped and left-skewed). It compares the medians of the samples. This is appropriate when the underlying distributions are not symmetric.

The independent t-test is more powerful when its assumptions are met, but when they are not, a non-parametric test like Mann-Whitney is has more power.

The null hypothesis of the Mann-Whitney U test is that there is no difference in the  medians of the two samples.

## Approach
Use histograms to assess the shape of the sample distributions to determine if a non-parametric test would be a good choice.

## Effect size and power
The `rank_biserial` function in the `effectsize` package can be used to compute the effect size of the Mann-Whitney U test. Generic effect size values for this test are

* Small: 0.1
* Medium: 0.3
* Large: 0.5

The `pwr.2p2n.test` function can be used to determine power.

```{r run Mann-Whitney U test}
wilcox.test(sum_gamerounds ~ version, 
            data = gamedata)
```
The *p*-value is large.

```{r Derive the effect size}
rank_biserial(sum_gamerounds ~ version, 
         data = gamedata)
```
The effect size is very small. This will require a lot of power to detect with confidence.

```{r run power analysis}
pwr.2p2n.test(h=0.00672,
              sig.level = 0.084,
              n1=1500,
              n2=1500)
```
The power of this particular test is quite small. There is a very high probability of a Type II error in this situation. With such a small effect size, the samples would have to be very large.

$$
\text{Probability of Type II error} = 1 - \text{Power}\\[0.25cm]
\text{Probability of Type II error} = 1 - 0.089 = 0.911
$$

```{r}
covid <- read_csv("data/covid-dat.csv")
```


```{r}
# Run an independent t-test assessing time on social media
t.test(Time.spent.on.social.media ~ Medium.for.online.class, data = covid, var.equal = TRUE)

cohens_d(Time.spent.on.social.media ~ Medium.for.online.class, data = covid)

# Run a power analysis of the t-test
pwr.t.test(n = 542, sig.level = 0.00652, d = 0.17, type = "two.sample")
```
Now, run a Mann-Whitney test on the same data.

```{r}
wilcox.test(Time.spent.on.social.media ~ Medium.for.online.class, data = covid)

rank_biserial(Time.spent.on.social.media ~ Medium.for.online.class, data = covid)
```
Run a power analysis on the Mann-Whitney test.

```{r}
pwr.2p2n.test(h=0.09,
              sig.level = 0.0061,
              n1=542,
              n2=542)
```

```{r}
covid |> ggplot(aes(x=Time.spent.on.social.media)) +
  geom_histogram() +
  facet_wrap(~Medium.for.online.class)
```

